---
title: "How to use the package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use the package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = F,
  warning = F,
  comment = "#>"
)
```

```{r setup}
library(NotAllOilTypesAreAlike)
```

The goal of this vignette is to approximate the production losses of one specific
scenario. For this example, we first estimate the elasticities of substitution 
for Germany and then approximate the production loss for the scenario where 
Germany loses access to Russian crude oil.

## Prepare the Dataset

```{r}
df <- merge_quality_quantity() %>%
  filter(`Reporting Country` == "Germany") %>%
  rename("Oil Field" = `Type of crude oil`) %>%
  mutate(date = as.numeric(substr(date, 1, 4))) %>%
  group_by(`Country of Origin`, `Oil Field`, `Reporting Country`,
            `date`, `Weight`, `Sulfur`) %>%
  summarize(`Volume (1000 bbl)` = sum(`Volume (1000 bbl)`, na.rm = T),
             `Total Value ($ 1000)` = sum(`Total Value ($ 1000)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(`CIF price (2) ($/bbl)` = `Total Value ($ 1000)` / `Volume (1000 bbl)`,
          date = as.Date(paste0(date, "-12-31")),
          type = paste(Weight, Sulfur))


```

To estimate the elasticity of substitution between countries of origin and 
between types we need to aggregate the data.

```{r}
# country level aggregation
df_country <- df %>%
  group_by(date, `Reporting Country`, `Country of Origin`) %>%
  summarize(`Volume (1000 bbl)` = sum(`Volume (1000 bbl)`, na.rm = T),
             `Total Value ($ 1000)` = sum(`Total Value ($ 1000)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(`CIF price (2) ($/bbl)` = `Total Value ($ 1000)` / `Volume (1000 bbl)`)

# oil type aggregation
df_type <- df %>%
  mutate(`Oil Type` = paste(Weight, Sulfur)) %>%
  group_by(date, `Reporting Country`, `Oil Type`) %>%
  summarize(`Volume (1000 bbl)` = sum(`Volume (1000 bbl)`, na.rm = T),
             `Total Value ($ 1000)` = sum(`Total Value ($ 1000)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(`CIF price (2) ($/bbl)` = `Total Value ($ 1000)` / `Volume (1000 bbl)`)
```

## Estimate the elasticity of substitution

In a first step we estimate the elasticity of substitution following the
standard Armington assumption. One country is now considered as one variety and
we ask how easy we can switch from one country to another.

The function `elasticity_feenstra1994` estimates many other parameters as well. 
However, we are only interested in `sigma`. 

```{r}
ela_country <- elasticity_feenstra1994(df     = df_country,
                                       t_name = "date",
                                       v_name = "Country of Origin",
                                       p_name = "CIF price (2) ($/bbl)",
                                       q_name = "Volume (1000 bbl)",
                                       freq   = "y")

ela_country
```

Now we treat one oil type as one variety. The elasticity is expected to decline
as it should be more difficult to switch between types than it is between 
countries.

```{r}
ela_type <- elasticity_feenstra1994(df     = df_type,
                                    t_name = "date",
                                    v_name = "Oil Type",
                                    p_name = "CIF price (2) ($/bbl)",
                                    q_name = "Volume (1000 bbl)",
                                    freq   = "y")

ela_type
```

The elasticity drops from `r round(ela_country$sigma, 1)` to 
`r round(ela_type$sigma, 1)`.

For the nested CES production function we also need within type elasticity 
estimates. Those can be estimated in the same way as above but now we use the 
raw data without any aggregation and filter for the relvant types.

```{r}
df$weight_sulfur = paste(df$Weight, df$Sulfur)
ela_within <- NULL

for(VARIETY in unique(df$weight_sulfur)){

  temp <- df %>% filter(weight_sulfur == VARIETY)

  out <- tryCatch({
    elasticity_feenstra1994(df     = temp,
                             t_name = "date",
                             v_name = "Oil Field",
                             p_name = "CIF price (2) ($/bbl)",
                             q_name = "Volume (1000 bbl)",
                             freq   = "y")
  }, error=function(e){ out <- as.data.frame(t(rep(NA, 12))); colnames(out) <- c("nObs", "nVarieties", "theta1", "theta2", "rho", "sigma", "theta1_se", "theta2_se", "rho_se", "sigma_se", "sigma_lb", "sigma_ub"); out })

  out$Variety <- VARIETY; out$Country <- "Germany"

  ela_within <- rbind(ela_within, out)

}

rownames(ela_within) <- unique(df$weight_sulfur)

ela_within
```

As we see in the above example, in some instances the elasticity cannot be 
estimated. This is the case when less than three varieties are available. In 
this case `sigma` cannot be estimated using Feenstra's method.

## Approximation of production loss

We first correct Russian Federation Oil Types. According to reports from Bruegel
almost all crude oil imported from Russia is Urals. Therefore, this seems to be 
a reasonable assumption to classify German imports from Russia as Urals crude
oil.

```{r}
df <- df %>%
  mutate(type = paste(Weight, Sulfur),
         `Oil Field` = ifelse(`Country of Origin` == "Russian Federation", "Urals", `Oil Field`),
         type = ifelse(`Country of Origin` == "Russian Federation", "medium sour", type))
```

Now we use our estimates from above to approximate production disruptions.

```{r}
df_prod_loss <- est_prod_loss(x          = df,
                              x_diff     = df %>% filter(`Country of Origin` != "Russian Federation"),
                              quantity   = "Volume (1000 bbl)",
                              type       = "type",
                              sigma_ces  = ela_country$sigma,
                              sigma_nces = mean(ela_within$sigma, na.rm = T),
                              gamma_nces = ela_type$sigma)

df_prod_loss
```

`est_prod_loss` estimates the loss in overall production from a CES and a nested
CES case for two different datasets. In the first we used all data available and 
to run our scenario we removed Russia from the second dataset. Now the impact of 
our results is the difference of those two estimates. Thus, taking crude oil 
quality into account, production disruptions are 
`r df_prod_loss[1,1] - df_prod_loss[1,2]` higher than using a standard CES production 
function.

# Session Info

```{r}
sessionInfo()
```

