---
title: "Not all oil types are alike - Replication of figures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{replicate-figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=F}
library(NotAllOilTypesAreAlike)
```

This vignette is a step-by-step guide to replicate each figure of the paper
"Not all oil types are alike". 

# Figure 2

To prepare the data for each panel, the function `merge_quality_quantity()` is 
employed. This custom function combines trade flow data with crude oil quality 
information. This function includes a parameter that allows for a variation of 
the Russian Urals import share. As highlighted in the paper the default is 
70% of all Russian imports are of type Urals. However, 70% is probably on the 
conservative side, as official sources say that Russian Urals is the most 
important type of crude oil imported from Russia, accounting for well over 80%. 

```{r}
df_raw <- merge_quality_quantity() %>%
  mutate(oil_type = paste(Weight, Sulfur)) %>%
  filter(!is.na(`Typical API gravity`),
         `Reporting Country` %in% get_ea_countries()) %>%
  na.omit()

df_country <- df_raw %>%
  group_by(`Country of Origin`, oil_type) %>%
  summarize(vol = sum(`Volume (1000 bbl)`),
            api = mean_api(`Typical API gravity`, 
                            weight = `Volume (1000 bbl)`),
            sul = mean(`Typical sulphur content`))

df_field <- df_raw %>%
  group_by(`Type of crude oil`, `Country of Origin`, oil_type) %>%
  summarize(vol = sum(`Volume (1000 bbl)`),
             api = mean_api(`Typical API gravity`, 
                             weight = `Volume (1000 bbl)`),
             sul = mean(`Typical sulphur content`))

df_type <- df_raw %>%
  group_by(oil_type) %>%
  summarize(vol = sum(`Volume (1000 bbl)`),
             api = mean_api(`Typical API gravity`, 
                             weight = `Volume (1000 bbl)`),
             sul = mean(`Typical sulphur content`))
```

We then create the panels of figure 2 using ggplot2.

```{r}
plot_field <- ggplot(df_field,
                      aes(x = api, y = sul, size = vol, 
                           color = oil_type, shape = oil_type)) +
  geom_point(alpha = 0.75) +
  scale_size(range = c(6, 30), guide = "none") +
  theme(legend.position = "none") +
  scale_shape_manual(values = c(17, 18, 19, 15, 8)) +
  annotate("segment", x = 35, y = 2.5, xend = 32.5, yend = 2.25,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", x=34,y=2.55, label = "Russian Urals", color = "#011936",
           angle = 0, hjust = 0.1, size = 3, fontface = "bold")

plot_country <- ggplot(df_country,
                        aes(x = api, y = sul, size = vol, 
                             color = oil_type, shape = oil_type)) +
  geom_point(alpha = 0.75) +
  scale_size(range = c(6, 30), guide = "none") +
  theme(legend.position = "none") +
  scale_shape_manual(values = c(17, 18, 19, 15, 8)) +
  annotate("segment", x = 31.5, y = 0.1, xend = 34.5, yend = 0.15,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", x=26,y=0.1, label = "Kazakhstan", color = "#011936",
           angle = 0, hjust = 0.1, size = 3, fontface = "bold")

plot_type <- ggplot(df_type,
                     aes(x = api, y = sul, size = vol, 
                          color = oil_type, shape = oil_type)) +
  geom_point(alpha = 0.75) +
  scale_size(range = c(6, 30), guide = "none") +
  theme(legend.position = "none") +
  scale_shape_manual(values = c(17, 18, 19, 15, 8)) +
  annotate("segment", x = 36, y = 2.7, xend = 33.4, yend = 2.5,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", x=34,y=2.85, label = "81% of which is\nRussian Urals", 
           color = "#011936", angle = 0, hjust = 0.1, size = 3, 
           fontface = "bold") +
  annotate("segment", x = 45, y = 0.715, xend = 42, yend = 0.5,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", x=41,y=0.84, label = "17% of which is\nKazakhstan crude",
           color = "#011936", angle = 0, hjust = 0.1, size = 3, 
           fontface = "bold")

size_min <- min(df_field$vol)
size_max <- max(df_type$vol)
```

Formatting is then carried out with user-defined functions:

```{r}
plot_field <- gx_theme(plot_field,
                        x_title = "API gravity [°]",
                        y1_title = "Sulfur content [%]",
                        legend_position = "none",
                        text_size = 12,
                        text_colour = "black") +
  scale_y_continuous(limits = c(0, 3.5)) +
  scale_x_continuous(limits = c(13, 51)) +
  scale_size(limits = c(size_min, size_max), range = c(4, 24)) +
  theme(axis.text.x = element_text(vjust = .1))

plot_country <- gx_theme(plot_country,
                          x_title = "API gravity [°]",
                          y1_title = "Sulfur content [%]",
                          legend_position = "none",
                          text_size = 12,
                          text_colour = "black") +
  scale_y_continuous(limits = c(0, 3.5)) +
  scale_x_continuous(limits = c(13, 51)) +
  scale_size(limits = c(size_min, size_max), range = c(4, 24)) +
  theme(axis.text.x = element_text(vjust = .1))

df_field$within_ela <- ifelse(df_field$oil_type == "light sweet", 0.8, 0.2)

plot_within_ela <- gx_theme(ggplot(df_field,
                                     aes(x = api, y = sul, size = vol, 
                                         color = oil_type, alpha = within_ela, 
                                         shape = oil_type)) +
                               geom_point() +
                               scale_size(range = c(6, 30), guide = "none") +
                               theme(legend.position = "none") +
                               scale_shape_manual(values = c(17, 18, 19, 15, 8)),
                             x_title = "API gravity [°]",
                             y1_title = "Sulfur content [%]",
                             legend_position = "none",
                             text_size = 12,
                             text_colour = "black") +
  scale_alpha(range = c(0.1, 0.75)) +
  scale_y_continuous(limits = c(0, 3.5)) +
  scale_x_continuous(limits = c(13, 51)) +
  scale_size(limits = c(size_min, size_max), range = c(4, 24)) +
  theme(axis.text.x = element_text(vjust = .1))

plot_type <- gx_theme(plot_type,
                       x_title = "API gravity [°]",
                       y1_title = "Sulfur content [%]",
                       legend_position = "none",
                       text_size = 12,
                       text_colour = "black") +
  scale_y_continuous(limits = c(0, 3.5)) +
  scale_x_continuous(limits = c(13, 51)) +
  scale_size(limits = c(size_min, size_max), range = c(4, 24)) +
  theme(axis.text.x = element_text(vjust = .1))

# create combined legend
p <- ggplot(df_country,
               aes(x = api, y = sul, color = oil_type, shape = oil_type)) +
    geom_point(alpha = 0.75) +
    scale_shape_manual(values = c(17, 18, 19, 15, 8)) +
    scale_color_manual(values = gx_colors()) +
    theme(legend.background = element_blank(),
          legend.position   = "bottom",
          legend.title      = element_blank(),
          legend.key        = element_blank(),
          legend.text       = element_text(size = 12))+
    guides(shape = guide_legend(override.aes = list(size = 5)))

# Extract the legend. Returns a gtable
leg <- ggpubr::get_legend(p)
```


```{r, fig.height=7, fig.width=7}
  # create and print plot
  ggpubr::ggarrange(plot_field, 
                    plot_country, 
                    plot_within_ela, 
                    plot_type, 
                    common.legend = T, legend="bottom", legend.grob = leg)
```


# Figure 3

In this figure we show the distribution of Euro Area imports for the two most 
common crude oil types. We do this for the whole sample period to avoid year 
specific variation. Dottedbars refer to medium-sour and striped bars refer to 
light-sweet crude oil imports. Russian imports of the two types are highlighted 
in red and blue, respectively.

Data preparation is straightforward and involves similar steps than in figure 2.

```{r}
df_ea <- merge_quality_quantity() %>%
  filter(`Reporting Country` %in% get_ea_countries())

df_ea <- df_ea %>%
  mutate(`Reporting Country` = "Euro Area",
         date = as.numeric(substr(date, 1, 4))) %>%
  group_by(`Country of Origin`, `Type of crude oil`, `Reporting Country`,
           `date`, `Weight`, `Sulfur`) %>%
  summarize(`Volume (1000 bbl)` = sum(`Volume (1000 bbl)`, na.rm = T),
            `Total Value ($ 1000)` = sum(`Total Value ($ 1000)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(`CIF price (2) ($/bbl)` = `Total Value ($ 1000)` / `Volume (1000 bbl)`,
          date = as.Date(paste0(date, "-12-31"))) %>%
  filter(`Total Value ($ 1000)` > 0, 
          `CIF price (2) ($/bbl)` > 0, 
          `Volume (1000 bbl)` > 0)

df <- df_ea %>%
  mutate(oil_type = paste(Weight, Sulfur)) %>%
  rename("country" = `Country of Origin`) %>%
  group_by(country, oil_type) %>%
  summarize(volume = sum(`Volume (1000 bbl)`)) %>%
  filter(volume > 5000) %>%
  filter(oil_type %in% c("light sweet", "medium sour")) %>%
  pivot_wider(names_from = oil_type, values_from = volume) %>%
  pivot_longer(-country) %>%
  mutate(value = ifelse(is.na(value), 0, value)) %>%
  mutate(value = value / 1000,
         country = ifelse(country == "Russian Federation", "Russia",
                          ifelse(country == "United Kingdom", "UK",
                                 ifelse(country == "United States", "USA", 
                                        country))))
```

Than we create the plot and use again our custom functions for formatting:

```{r, fig.height=7, fig.width=7}
plot <- ggplot(df, aes(x = country, y = value)) +
  geom_col_pattern(aes(pattern = name),
                   position        = "dodge",
                   fill            = 'white',
                   colour          = 'black',
                   pattern_spacing = 0.015) +
  geom_col_pattern(data            = filter(df, country == "Russia"),
                   mapping         = aes(fill = name, pattern = name),
                   position        = "dodge",
                   pattern_spacing = 0.015)

gx_theme(plot,
         text_size   = 12,
         text_colour = "black",
         y1_title    = "Crude oil imports in million barrels") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1.05, vjust = 0.25))
```

# Figure 4

In this figure we show European crude oil imports from the United States and 
Russia before and after the start of the EU crude oil embargo. Panel a) Time
series of imported volumes from Russia and the U.S. into the EU. Panel b) 
Chemical composition of crude oils imported from Russia [blue squares] in the 
year before the sanctions and chemical composition of U.S. imports [red bubbles] 
in the first year of the oil embargo. The size of the symbols refers to the 
volume of crude oil imports.

For this analysis we need a different dataset. We use data provide by Eurostat.
At the time of writing this data contained crude oil imports into the European
Union from January 2020 to October 2023 on a monthly basis. As this dataset is 
quite large and thus not suitable for keeping it in the package, I converted the
.csv-file to .RDS which is much more memory efficient. It is still the raw data
but now can be easily accessed via the package. 

The data preparation is again straightforward. The only thing which is done in 
the following code block is reshaping the data and renaming the variables such
that the code is easy to read. 

```{r}
df <- prepare_eurostat()
```

To make it easier to follow we will separate the two panels and create one plot
for each panel.

## Panel a

We apply a country filtering as we are only interested in imports from Russia 
and the United States. Than we aggregate across different oil types to obtain 
the total volume imported from each of the two countries.

```{r, fig.height=7, fig.width=7}
df_plot <- df %>%
  filter(repCountryIso == "EU", sourceCountryIso %in% c("RU", "US")) %>%
  mutate(sourceCountryIso = ifelse(sourceCountryIso == "RU",
                                   "Crude oil imports from Russia",
                                   "Crude oil imports from the U.S.")) %>%
  group_by(month, sourceCountryIso) %>%
  summarize(volume = sum(volume),
            value = sum(value)) %>%
  mutate(avgPriceBBl = value / volume)

# continue missing RU imports with zeros (as they are true zeros)
df_plot_vol <- df_plot %>%
  select(month, sourceCountryIso, volume) %>%
  pivot_wider(names_from = month, values_from = volume)

df_plot_vol[is.na(df_plot_vol)] <- 0

df_plot_vol <- df_plot_vol %>%
  pivot_longer(-sourceCountryIso, names_to = "month", values_to = "volume") %>%
  mutate(month = as.Date(month),
         volume = volume/1000)

plot_vol <- ggplot(df_plot_vol, aes(x = month, y = volume, color = sourceCountryIso)) +
  geom_point(size = 1.2) +
  geom_line(size = 1.1) +
  geom_vline(xintercept = as.Date("2022-12-05"), color = "black", size = 1.2) +
  annotate("text", x=as.Date("2022-12-05"),y=63.5, label = "EU embargo\ntakes effect", color = "black",
           hjust = -0.025, size = 5, fontface = "bold")

gx_theme(plot_vol,
         y1_title = "Crude oil imports in million barrels",
         text_size = 12,
         legend_rows = 2,
         text_colour = "black") +
  scale_x_date(limit = c(as.Date("2020-01-01"), as.Date("2023-12-30")))
```

## Panel b

In this panel we compare the quality of crude oil imports from Russia and the
United States. We focus on the year before and after the embargo comes into
forces. Here we should not that the crude oil embargo was implemented on 
December 5th, 2022. In the quality comparison we use November 2022 as the start
of the sanctions which does not influence the results. It is just that at the 
time of writing, the last observation available was October 2023. When we use 
November, we can use 12 months for both periods (Russian imports before the 
sanctions and U.S. imports after the sanctions).

The only thing we need to take care of is the aggregation of oil quality. Here 
we apply our custom function `mean_api()` as we can not simply take the average
of API gravity. We first need to convert it to specific gravity before we can 
take means and convert back.


```{r, fig.height=7, fig.width=7}
df_oil_quality_ru <- df %>%
  filter(repCountryIso == "EU", sourceCountryIso %in% c("RU"),
         month <= as.Date("2022-10-31"),
         month > as.Date("2021-10-31")) %>%
  mutate(volShare = volume/sum(volume),
         meanApi = mean_api(api, volShare),
         meanSul = sum(sul*volShare)) %>%
  select(meanApi, meanSul) %>%
  unique()

df_oil_quality_us <- df %>%
  filter(repCountryIso == "EU", sourceCountryIso %in% c("US"),
         month > as.Date("2022-10-31")) %>%
  mutate(volShare = volume/sum(volume),
         meanApi = mean_api(api, volShare),
         meanSul = sum(sul*volShare)) %>%
  select(meanApi, meanSul) %>%
  unique()

df_plot_qual <- df %>%
  filter(repCountryIso == "EU", sourceCountryIso %in% c("RU", "US"),
         ((sourceCountryIso == "RU" & month <= as.Date("2022-10-31")) |
            (sourceCountryIso == "US" & month > as.Date("2022-10-31")))) %>%
  mutate(sulType = ifelse(sul < 1, "sweet", "sour"),
         apiType = ifelse(api < 22.3, "heavy", ifelse(api > 31.1, "light", "medium")),
         oilType = paste(apiType, sulType))

plot_quality_us_ru <- df_plot_qual %>%
  ggplot(aes(x = api, y = sul,
             size  = volume,
             color = sourceCountryIso,
             shape = sourceCountryIso)) +
  geom_point(alpha = 0.75 ) +
  scale_size(range = c(2, 14), guide = "none" ) +
  scale_shape_manual(values = c(15, 19))

gx_theme(plot_quality_us_ru,
          x_title     = "API gravity [°]",
          y1_title    = "Sulfur content [%]",
          legend_rows = 2,
          text_size   = 12,
          text_colour = "black") +
  theme(axis.text.x   = element_text(vjust = .1)) +
  scale_color_manual(values = c(gx_colors()[1:2]),
    labels = c("RU" = "Pre-embargo crude oil imports from Russia",
               "US" = "Post-embargo crude oil imports from the U.S.")) +
  scale_shape_manual(
    values = c(15,19),
    labels = c("RU" = "Pre-embargo crude oil imports from Russia",
               "US" = "Post-embargo crude oil imports from the U.S.")) +
  guides(color = guide_legend(nrow = 2),
         shape = guide_legend(nrow = 2, override.aes = list(size = 4)))

```

# Session Info

```{r}
sessionInfo()
```

